\documentclass[a4paper]{article}
\usepackage{fullpage}
\usepackage{hyperref}
\usepackage{alltt}

\title{Purity inference}
\author{Jacques Garrigue}

\let\th\vdash
\let\E\Sigma
\let\G\Gamma
\let\ra\rightarrow
\let\tv\alpha
\let\t\tau
\let\s\sigma
\def\p{\mathsf{p}}
\def\i{\mathsf{i}}
\newcommand\lfrac[2]{\frac{#1\hfill}{#2\hfill}}
\newcommand\mlet[1]{\textsf{let}~#1~\textsf{in}~}
\newcommand\mletrec[1]{\textsf{let\,rec}~#1~\textsf{in}~}
\newcommand\mfun[1]{\textsf{fun}~#1 \ra}
\let\type\textit
\newcommand\cmt[1]{\hfill{(* #1 *)}}

\begin{document}
\maketitle

The value restriction, even in its relaxed form, makes us lose
polymorphism even in cases where all functions are pure.
This is especially galling for combinator libraries, where code is
built through function applications.

Can we do better than that?
The value restriction is only a weakened form of the original
strong/weak type variable distinction present in Standard ML 90.
It is a radical one: we only keep weak variables, deeming strong ones
not modular enough.

Reintroducing strong variables would certainly be theoretically be
fine, but there are practical reasons for their demotion.
Here we will argue for a coarser grain approach: distinguish between
pure and impure bindings in the environment.
Our soundness criterion is as follows: a binding can be marked as pure
if all its polymorphic type variables would be strong using the
original SML 90 weak/strong discipline.
This also shows the limitation of this approach: while SML 90 allowed
mixing strong and weak type variables in the same polymorphic type, we
allow only one category.

\[ s ::= \p \mid \i
\hspace{8ex}
\p \land s = s
\hspace{8ex}
\i \land s = \i
\]

\[
\E \th x : s  \quad (x:s)\in E
\hspace{8ex}
\lfrac{\E, x : \p \th M : s}
     {\E \th \mfun x M : s}
\hspace{8ex}
\lfrac{\E \th M_1 : s_1 \quad \E \th M_2 : s_2}
{\E \th M_1 \, M_2 : s_1 \land s_2}
\]

\[
\lfrac{\E \th M_1 : s_1 \quad \dots \quad \E \th M_n : s_n}
{\E \th (M_1, \dots, M_n) : s_1 \land \dots \land s_n}
\hspace{8ex}
\lfrac{\E \th M_1 : s_1 \quad \E \th M_2 : s_2}
{\E \th (M_1 ; M_2) : s_2}
\]

\[
\lfrac{\E \th M_1 : s_1 \quad \E, x : s_1 \th M_2 : s_2}
{\E \th \mlet{x=M_1} M_2 : s_2}
\hspace{8ex}
\lfrac{\E, x : \p \th M_1 : s_1 \quad \E, x : s_1 \th M_2 : s_2}
{\E \th \mletrec{x=M_1} M_2 : s_2}
\]
\[
\lfrac{\E, x : \i \th M_1 : s_1 \quad \E, x : s_1 \th M_2 : s_2}
{\E \th \mletrec{x:\forall\tv.\sigma=M_1} M_2 : s_2}
\hspace{8ex}
\lfrac{\E \th r : s_1 \quad \th_{\sf lab} l : s_2}
{\E \th r.l : s_1 \land s_2}
\]

\[
\th_{\sf lab} l :
\left\{\begin{array}{ll}
\i & \mbox{when $l$ is a polymorphic field} \\
\p & \mbox{otherwise} \\
\end{array}\right.
\quad
\begin{array}{l}
  \mbox{Polymorphic fields inside patterns} \\
  \mbox{also make the expression impure.}
\end{array}
\]

\[
\lfrac{\E \th M_i : s_i \quad \th_{\sf mut} l_i : s'_i}
{\E \th \{l_1 = M_1; \dots; l_n = M_n\} : \bigwedge_i s_i \land s'_i}
\hspace{5ex}
\th_{\sf mut} l :
\left\{\begin{array}{ll}
\i & \mbox{when $l$ is a mutable field} \\
\p & \mbox{otherwise} \\
\end{array}\right.
\]

\[
\lfrac{\G;\E \th M_1 : \t_1  \quad \E \th M_1 : \p \quad
  \G, x : \forall\vec\tv.\t_1; \E,x : \p \th M_2 : \t_2}
{\G;\E \th \mlet{x = M_1} M_2 : \t_2}~ \vec\tv \cap \textsf{fv}(\G) = \emptyset
\]

This solution is implemented by my {\tt purity-attribute} branch:
\url{https://github.com/garrigue/ocaml/tree/purity-attribute}.
Typed examples are at:
\url{https://github.com/garrigue/ocaml/blob/purity-attribute/testsuite/tests/typing-poly/purity.ml}.

One limitation of the current implementation is that simultaneous
let-definitions (using {\tt and}) are type as though they were tuples
in the above rules.

\paragraph{Examples}
\begin{alltt}
(* Basic *)

let some x = Some x;; \cmt{constructor functions can be pure}
\type{val some : 'a -> 'a option [@@pure]}
let a = some (fun x -> x);; \cmt{pure applications can be generalized}
\type{val a : ('a -> 'a) option [@@pure]}

let r = ref;; \cmt{impure things are not pure!}
\type{val r : 'a -> 'a ref}
let c = r (fun x -> x);; \cmt{relaxed value restriction still applies}
\type{val c : ('_weak1 -> '_weak1) ref}
let c' = {contents=fun x -> x};;
\type{val c' : ('_weak2 -> '_weak2) ref}

(* Modules *)

module type T = sig val some : 'a -> 'a option [@@pure] end;;

module M : T = struct let some x = Some x end;;
\type{module M : T}

module M' : T = struct let some x = let r = ref x in Some !r end;;
\type{Error: Signature mismatch:
       Values do not match:
         val some : 'a -> 'a option
       is not included in
         val some : 'a -> 'a option [@@pure]}

(* Beware of polymorphism *)
type mkref = \{mkref: 'a. 'a -> 'a ref\};; \cmt{polymorphic field}

let f x = let r = x.mkref [] in fun y -> r := [y]; List.hd !r;;
\type{val f : mkref -> 'a -> 'a} \cmt{impure field access}

let f \{mkref\} = let r = mkref [] in fun y -> r := [y]; List.hd !r;;
\type{val f : mkref -> 'a -> 'a} \cmt{impure pattern}

(* Beware of polymorphic recursion *)
let rec f x = let r = mkref x [] in fun y -> r := [y]; List.hd !r
and mkref : 'a. mkref -> 'a -> 'a ref = fun x -> x.mkref;;
\type{val f : mkref -> 'a -> 'a} \cmt{call to mkref is impure}
\type{val mkref : mkref -> 'a -> 'a ref}

let rec id : 'a. 'a -> 'a = fun x -> id x;;
\type{val id : 'a -> 'a} \cmt{impure due to polymorphic recursion}

let rec id : 'a. 'a -> 'a = fun x -> x;;
\type{val id : 'a -> 'a [@@pure]} \cmt{pure since no recursive call}
\end{alltt}

\end{document}
